---
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    highlight: zenburn
    theme: flatly
---

```{r load-data}
library(dplyr)
library(ggplot2)
infn = "../data/human.csv"
metadata = read.table(infn, sep=",", header=TRUE)
metadata$condition = ifelse(grepl("Healthy", metadata$What), "HC",
                     ifelse(grepl("NASH", metadata$What), "OSA",
                     ifelse(grepl("Steat", metadata$What), "SS", "NAFLD")))
metadata$condition = as.factor(metadata$condition)
metadata$spiked = ifelse(grepl("Spike", metadata$Spike.In), "spike", "nospike")
metadata$column = ifelse(grepl("ymo", metadata$Column), "zymo",
                  ifelse(grepl("None", metadata$Column), "none", "other"))
metadata$id = paste(metadata$Full.Sample.Name, metadata$column, metadata$spiked, metadata$condition, sep="_")
```

```{r load-files}
library(NanoStringQCPro)
rccFiles = paste("../data/rcc/", metadata[,2], sep="")
eset = newRccSet(rccFiles=rccFiles)
colnames(eset) = metadata$id
```

## Scale miRNA with known crosstalk
A small set of miRNA have known inflated values; Nanostring has a heuristic to
correct for the inflation via the formula using some supplied constants. Here
we implement that fix to correct the expression of those miRNA.

```{r add-scaledata}
correct_miRNA = function(eset) {
  require(dplyr)
  fdat = fData(eset)
  fdat = fdat %>%
    tidyr::separate(GeneName, c("Name", "scalefactor"), sep='\\|',
                    fill="right", remove=FALSE) %>%
    dplyr::mutate(scalefactor=ifelse(is.na(scalefactor), 0, scalefactor))
  sf = as.numeric(fdat$scalefactor)
  posa = as.numeric(exprs(eset)["Positive_POS_A_ERCC_00117.1",])
  scales = t(replicate(length(sf), posa))
  scales = scales * sf
  scaled = exprs(eset) - scales
  scaled[scaled<0] = 0
  return(round(scaled))
}

plot_corrected_miRNA = function(scaled, eset) {
  require(reshape)
  require(ggplot2)
  counts = exprs(eset)
  rownames(counts) = rownames(scaled)
  is_scaled = rowSums(abs(counts - scaled)) > 0
  counts = melt(as.matrix(counts[is_scaled,]))
  colnames(counts) = c("gene", "sample", "value")
  counts$scaled = "no"
  scaled = melt(as.matrix(scaled[is_scaled,]))
  colnames(scaled) = c("gene", "sample", "value")
  scaled$scaled = "yes"
  all = rbind(counts, scaled)
  library(cowplot)
  ggplot(all, aes(sample, value, color=scaled)) + facet_wrap(~gene) +
    geom_point(size=0.5) +
    theme(axis.title.x=element_blank(),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank(),
          strip.text.x=element_text(size=8)) +
    guides(colour = guide_legend(override.aes = list(size=6))) + ylab("") +
    ggtitle("miRNA hybridization crosstalk correction")
  }
```

```{r scale-counts}
scaled = correct_miRNA(eset)
plot_corrected_miRNA(scaled, eset)
counts = as.data.frame(scaled)
colnames(counts) = metadata$id
```

## Imaging QC
This percentage should be super high, close to 100%. If there isn't that usually means
the cartridge needs to be rescanned. If it is < 3 weeks from the scan, rescanning
should be okay. Nanostring folks call < 75% a failure. These all look fine.

```{r imaging-qc}
library(cowplot)
pdat = pData(eset) %>%
  tibble::rownames_to_column() %>%
  left_join(metadata, by=c("rowname"="id"))
pdat$pcounted = pdat$FovCounted/pdat$FovCount * 100
ggplot(pdat, aes(rowname, pcounted)) + geom_point() +
    theme(axis.text.x=element_blank(),
          axis.ticks.x=element_blank(),
          strip.text.x=element_text(size=8)) +
  scale_y_continuous(expand = c(0,0)) +
  expand_limits(y = c(0,1.05 * max(pdat$pcounted))) +
  ylab("percentage of FOV counted") + xlab("sample") +
  geom_hline(yintercept=75, color="red")
```

## Binding density
Binding density looks ok.
```{r binding-density}
ggplot(pdat, aes(rowname, BindingDensity)) + geom_point() +
    theme(axis.text.x=element_blank(),
          axis.ticks.x=element_blank(),
          strip.text.x=element_text(size=8)) +
  scale_y_continuous(expand = c(0,0)) +
  expand_limits(y = c(0,1.05 * max(pdat$BindingDensity))) +
  ylab("Binding density") + xlab("sample") +
  geom_hline(yintercept=0.05, color="red") +
  geom_hline(yintercept=2.25, color="red")
```

## Total counts vs miRNA detected
Here I plotted the total counts vs the number of miRNA detected, where
detected is it has counts > 30. There is a pretty big spread among the samples
in terms of the miRNA detected. The Zymo columns seem to all have a low number of
miRNA detected.

```{r non-controlcounts}
endocounts = counts[grepl("Endo", rownames(counts)),]
cdf = data.frame(total=colSums(counts), detected=colSums(counts > 30))
rownames(cdf) = colnames(counts)
cdf$id = rownames(cdf)
cdf = cdf %>% left_join(metadata, by="id")
ggplot(cdf, aes(total, detected, color=column, label=condition)) +
  geom_text()
```

```{r positive}
library(ggplot2)
library(dplyr)
library(cowplot)
is_positive = function(column) {
  return(grepl("Pos", column))
}
is_negative = function(column) {
  return(grepl("Neg", column))
}
is_spikein = function(column) {
  return(grepl("Spike", column))
}
is_ligation = function(column) {
  return(grepl("Ligati", column))
}
is_housekeeping = function(column) {
  return(grepl("Housekee", column))
}
is_prior = function(column) {
  return(grepl("miR-159", column) | grepl("miR-248", column) |
         grepl("miR-254", column))
}

extract_pred = function(counts, predicate) {
  toplot = counts[predicate(rownames(counts)),] %>%
    tibble::rownames_to_column() %>%
    tidyr::gather("sample", "count", -rowname)
  colnames(toplot) = c("spot", "sample", "count")
  toplot = toplot %>% left_join(metadata, by=c("sample"="id"))
  return(toplot)
}
spotbarplot = function(toplot) {
  ggplot(toplot,
        aes(sample, count)) + geom_bar(stat='identity') +
    facet_wrap(~spot) +
    theme(axis.text.x = element_blank(),
          text = element_text(size=8))
}
spotboxplot = function(toplot) {
  ggplot(toplot,
        aes(linehypo, count)) + geom_boxplot() +
    facet_wrap(~spot) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
}
```

## Positive controls
Below we look at the R^2 correlation between the expected positive
control concentrations and the observed concentrations for each
sample.

A set of samples have a lower correlation than the other samples.

```{r pos-r2}
spotbarplot(extract_pred(counts, is_positive))
pcdf = data.frame(concentration=log2(c(128, 32, 8, 2, 0.5, 0.125)),
                  GeneName=paste("POS", c("A", "B", "C", "D", "E", "F"), sep="_"))
pccounts = subset(exprs(eset), grepl("Positive_POS", rownames(exprs(eset))))
pccounts = pccounts[sort(rownames(pccounts)),]
rownames(pccounts) = pcdf$GeneName
corsamples = data.frame(correlation=apply(pccounts, 2,
                                          function(x) cor(x, pcdf$concentration)),
                        sample=colnames(pccounts)) %>%
  left_join(metadata, by=c("sample"="id"))
ggplot(corsamples, aes(sample, correlation)) + geom_point() +
    theme(axis.text.x=element_blank(),
          axis.ticks.x=element_blank(),
          strip.text.x=element_text(size=8)) +
  scale_y_continuous(expand = c(0,0)) +
  expand_limits(y = c(0,1.05 * max(corsamples$correlation))) +
  ylab("positive control correlation") +
  xlab("sample")
```

Those are all Zymo columns. 8 of the 10 Zymo columns have this problem. If it
looks like they have other issues, it might be best to just exclude them.

```{r poor-samples}
subset(corsamples, correlation < 0.80)$sample
```

## Negative controls
We can see some samples have a higher negative control count than the
other samples.
```{r negative-controls}
spotbarplot(extract_pred(counts, is_negative))
```

```{r high-negative}
knitr::kable(subset(extract_pred(counts, is_negative), count > 50))
```

## Noise floor cutoff
We'll normalize the libraries and look at the negative control expression and
then come up with a cutoff that is above the noise floor for the libraries.
It looks like 30 is a reasonable noise floor to use.

```{r noise-floor}
drop_unusable = function(counts) {
  drop = is_spikein(rownames(counts))
  drop = drop | is_positive(rownames(counts))
  drop = drop | is_housekeeping(rownames(counts))
  drop = drop | is_ligation(rownames(counts))
  keep = counts[!drop,]
  keep = keep[, !grepl("Blank", colnames(keep))]
  return(keep)
}
library(DESeq2)
dds = DESeqDataSetFromMatrix(drop_unusable(counts), colData=metadata,
                             design=~spiked+column+condition)
dds = estimateSizeFactors(dds)
ncounts = data.frame(counts(dds, normalized=TRUE))
ggplot(extract_pred(ncounts, is_negative), aes(count)) + geom_histogram()
nfloor = 30
```

## SpikeIn
We don't have many spike in sets in this sample either, so we can't really
use the spike ins.
```{r spikein}
spotbarplot(extract_pred(counts, is_spikein))
```

```{r spike-in-samples}
knitr::kable(subset(extract_pred(counts, is_spikein), count > 100))
```

## Ligation
Some of the ligation controls are drastically different between samples.

```{r ligation}
spotbarplot(extract_pred(counts, is_ligation))
```

## Ligation control R^2
Here we look at R^2 for the ligation controls as well. It looks like one
sample looks bad. It is a Zymo column.

```{r ligation-r2}
pcdf = data.frame(concentration=log2(c(128, 32, 8)),
                  GeneName=paste("POS_", c("A", "B", "C"), sep="_"))
pccounts = subset(exprs(eset), grepl("Ligation_LIG_POS", rownames(exprs(eset))))
pccounts = pccounts[sort(rownames(pccounts)),]
rownames(pccounts) = pcdf$GeneName
corsamples = data.frame(correlation=apply(pccounts, 2,
                                          function(x) cor(x, pcdf$concentration)),
                        sample=colnames(pccounts)) %>%
  left_join(metadata, by=c("sample"="id"))
ggplot(corsamples, aes(sample, correlation, color=column)) + geom_point() +
    theme(axis.text.x=element_blank(),
          axis.ticks.x=element_blank(),
          strip.text.x=element_text(size=8)) +
  scale_y_continuous(expand = c(0,0)) +
  expand_limits(y = c(0,1.05 * max(corsamples$correlation))) +
  ylab("ligation control correlation") +
  xlab("sample")
```

## Ligation efficiency
Here we calculated ligation efficiency for the samples. This is the
log of LIG_POS_A(i) / mean(LIG_POS_A) for each sample i. It doesn't matter
which direction this is in, it will get corrected for in the model. Again
we can see most of the Zymo columns are outliers in this.

```{r ligation-efficiency}
lig_posa = as.numeric(counts[grepl("Ligation_LIG_POS_A", rownames(counts)),])
metadata$lignorm = log2(lig_posa / mean(lig_posa))
ggplot(metadata, aes(id, lignorm, color=column)) + geom_point() +
    theme(axis.text.x=element_blank(),
          axis.ticks.x=element_blank(),
          strip.text.x=element_text(size=8)) +
  ylab("ligation factor") +
  xlab("sample")
```

## Housekeeping
Some housekeeping genes are expressed more highly in some samples. Not sure
what we are supposed to do with this information, if there are any suggestions
we could easily implement it.
```{r housekeeping}
spotbarplot(extract_pred(counts, is_housekeeping))
```

## Drop control miRNA
```{r drop-control}
drop_unusable = function(counts) {
  drop = is_spikein(rownames(counts))
  drop = drop | is_positive(rownames(counts))
  drop = drop | is_housekeeping(rownames(counts))
  drop = drop | is_ligation(rownames(counts))
  keep = counts[!drop,]
  keep = keep[, !grepl("Blank", colnames(keep))]
  return(keep)}
counts = drop_unusable(counts)
metadata = subset(metadata, id %in% colnames(counts))
```

## Does normalization for ligation effiency buy us anything?
Here we fit two models. The first model we fit the full model,
including the ligation normalization term in the model.
~month+column+line+hypoxia+lignorm. Then we fit a reduced model without
the lignorm term to answer the question 'is there miRNA that are affected by the
ligation efficiency?'

Yes, it looks like even after normalization, if we fit a model with just
the ligation efficiency that there are miRNA affected. So we should correct
for that. Here you can see only a subset of the miRNA are affected; if we
do it like this instead of just scaling the counts, we can correct for
the specific miRNA that have counts that are correlated with the ligation
efficiency.

```{r liga}
full = ~spiked+column+lignorm+condition
reduced = ~spiked+column+condition
dds = DESeqDataSetFromMatrix(counts, colData=metadata,
                             design=full)
dds = DESeq(dds, full=full, reduced=reduced, test="LRT")
res = results(dds)
plotMA(res)
res = data.frame(res) %>% tibble::rownames_to_column() %>%
  arrange(padj) %>% filter(padj < 0.05)
knitr::kable(subset(res, padj < 0.05))
```

## PCA

Here we can see the choice of column is what separates out the samples, it is
going to be hard to call differences because the column type is confounded
with the sample type for the most part. We only have one observation of
non-zymo columns for the OSA and SS samples.

```{r de-setup}
dds = DESeqDataSetFromMatrix(counts, colData=metadata, design=full)
vst = varianceStabilizingTransformation(dds)
pca_loadings = function(object, ntop=500) {
  rv <- matrixStats::rowVars(assay(object))
  select <- order(rv, decreasing = TRUE)[seq_len(min(ntop,
      length(rv)))]
  pca <- prcomp(t(assay(object)[select, ]))
  percentVar <- pca$sdev^2/sum(pca$sdev^2)
  names(percentVar) = colnames(pca$x)
  pca$percentVar = percentVar
  return(pca)}
pc = pca_loadings(vst)
comps = data.frame(pc$x)
comps$Name = rownames(comps)
library(dplyr)
comps = comps %>% left_join(metadata, by=c("Name"="id"))
pca_plot = function(comps, nc1, nc2, colorby) {
   c1str = paste0("PC", nc1)
   c2str = paste0("PC", nc2)
  ggplot(comps, aes_string(c1str, c2str, color=colorby)) +
    geom_point() + theme_bw() +
    xlab(paste0(c1str, ": ", round(pc$percentVar[nc1] * 100), "% variance")) +
    ylab(paste0(c2str, ": ", round(pc$percentVar[nc2] * 100), "% variance"))
  }
pca_plot(comps, 1, 2, "condition")
pca_plot(comps, 1, 2, "column")
```

It isn't clear what is separating out the samples on the lower left.

## Differential expression
```{r mirna-dispersion}
write_results = function(res, fname) {
  res = data.frame(res) %>% tibble::rownames_to_column() %>%
    arrange(pvalue)
  write.table(res, file=fname, quote=FALSE, col.names=TRUE,
              row.names=FALSE, sep=",")
}
dds = DESeq(dds, fitType='local')
plotDispEsts(dds)
```

## OSA/NASH vs HC
```{r osa-vs-hc}
res = results(dds, contrast=c("condition", "OSA", "HC"))
plotMA(res)
write_results(res, "OSA-vs-HC.csv")
```

## OSA/NASH vs SS
```{r osa-vs-ss}
res = results(dds, contrast=c("condition", "OSA", "SS"))
plotMA(res)
write_results(res, "OSA-vs-SS.csv")
```

## SS vs HC
```{r ss-vs-hc}
res = results(dds, contrast=c("condition", "SS", "HC"))
plotMA(res)
write_results(res, "SS-vs-HC.csv")
```

## Output tables
[OSA vs HC](OSA-vs-HC.csv)

[OSA vs SS](OSA-vs-SS.csv)

[SS vs HC](SS-vs-HC.csv)
